---
title: "Maine Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maine Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(meschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6")
```

```{r fetch-data, cache = TRUE, error = TRUE}
# Get available years
years <- get_available_years()
if (is.list(years)) {
  max_year <- years$max_year
  min_year <- years$min_year
} else {
  max_year <- max(years)
  min_year <- min(years)
}

# Attempt to fetch data - may fail if data source is unavailable
# (Maine DOE has moved to QuickSight dashboards with no direct download)
data_available <- tryCatch({
  enr <- fetch_enr_multi(max(min_year, max_year - 9):max_year)
  enr_current <- fetch_enr(max_year)
  TRUE
}, error = function(e) {
  message("Note: Maine DOE enrollment data is currently not available for automated download.")
  message("Visualizations in this vignette require manually downloaded data.")
  message("Use get_download_instructions() for guidance.")
  FALSE
})

# If data not available, skip remaining chunks
if (!data_available) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## 1. Maine is one of the whitest states in America

Over 90% of Maine students are white - the highest percentage of any state east of the Mississippi.

```{r demographics}
demo <- enr_current %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  mutate(subgroup_label = reorder(subgroup, -pct))

ggplot(demo, aes(x = subgroup_label, y = pct * 100)) +
  geom_col(fill = colors["total"]) +
  labs(title = "Maine is One of the Whitest States",
       subtitle = "Over 90% white - highest east of the Mississippi",
       x = "", y = "Percent of Students") +
  theme_readme()
```

## 2. Portland is diversifying rapidly

Portland Public Schools looks nothing like the rest of Maine - over 40% students of color and growing.

```{r portland-diversity}
portland <- enr %>%
  filter(is_district, grepl("Portland", district_name, ignore.case = TRUE),
         grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian"))

ggplot(portland, aes(x = end_year, y = pct * 100, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = colors,
                     labels = c("Asian", "Black", "Hispanic", "White")) +
  labs(title = "Portland is Diversifying Rapidly",
       subtitle = "Over 40% students of color - unlike rest of Maine",
       x = "School Year", y = "Percent", color = "") +
  theme_readme()
```

## 3. Rural Maine is losing students fast

Small SAUs in Aroostook County (the County) have lost 20-30% of enrollment since 2016.

```{r rural-decline}
state_trend <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment")

ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Rural Maine Losing Students",
       subtitle = "Aroostook County SAUs down 20-30% since 2016",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 4. Lewiston: Somali refugees transform a mill town

Lewiston has become one of the most diverse cities in New England thanks to refugee resettlement.

```{r lewiston-refugees}
lewiston <- enr %>%
  filter(is_district, grepl("Lewiston", district_name, ignore.case = TRUE),
         grade_level == "TOTAL", subgroup == "black")

ggplot(lewiston, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["black"]) +
  geom_point(size = 3, color = colors["black"]) +
  labs(title = "Lewiston: Somali Refugees Transform a Mill Town",
       subtitle = "Black student population growing as refugees resettle",
       x = "School Year", y = "Percent Black Students") +
  theme_readme()
```

## 5. COVID's kindergarten dip

Maine lost nearly 10% of kindergartners in 2021 - families kept kids home an extra year.

```{r covid-k}
k_trend <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "06", "12")) %>%
  mutate(grade_label = case_when(
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "06" ~ "Grade 6",
    grade_level == "12" ~ "Grade 12"
  ))

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID's Kindergarten Dip",
       subtitle = "Maine lost nearly 10% of kindergartners in 2021",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 6. Southern Maine is growing

Cumberland and York counties are gaining students while the rest of the state declines.

```{r southern-growth}
southern <- c("Portland", "South Portland", "Scarborough", "Falmouth")
south_trend <- enr %>%
  filter(is_district, grepl(paste(southern, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

ggplot(south_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Southern Maine Growing",
       subtitle = "Cumberland and York counties gaining while rest declines",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 7. Many SAUs have fewer than 500 students

Maine has more tiny school districts than almost any other state - some with under 100 students.

```{r district-sizes}
sizes <- enr_current %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(size = case_when(
    n_students < 100 ~ "Under 100",
    n_students < 500 ~ "100-499",
    n_students < 1000 ~ "500-999",
    n_students < 5000 ~ "1,000-4,999",
    TRUE ~ "5,000+"
  )) %>%
  group_by(size) %>%
  summarize(n_districts = n(), .groups = "drop")

ggplot(sizes, aes(x = size, y = n_districts)) +
  geom_col(fill = colors["total"]) +
  labs(title = "Many SAUs Have Fewer Than 500 Students",
       subtitle = "Maine's fragmented district structure",
       x = "SAU Size", y = "Number of SAUs") +
  theme_readme()
```

## 8. Bangor is stable in a declining region

Bangor maintains steady enrollment while surrounding Penobscot County shrinks.

```{r bangor-stable}
bangor <- enr %>%
  filter(is_district, grepl("Bangor", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(bangor, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Bangor: Stable in a Declining Region",
       subtitle = "Maintains enrollment while Penobscot County shrinks",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 9. English learners concentrated in a few districts

Portland, Lewiston, and a handful of other districts serve the vast majority of Maine's EL students.

```{r el-concentration}
el <- enr_current %>%
  filter(is_district, subgroup == "lep", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, n_students))

ggplot(el, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "English Learners Concentrated in Few Districts",
       subtitle = "Portland and Lewiston serve the vast majority",
       x = "", y = "English Learner Students") +
  theme_readme()
```

## 10. The graying of Maine shows in the schools

Maine's population is aging faster than any other state - and schools are feeling it first.

```{r statewide-trend}
ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "The Graying of Maine Shows in Schools",
       subtitle = "Population aging faster than any other state",
       x = "School Year", y = "Students") +
  theme_readme()
```
